---
title: "Data analysis on SQLite"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show 
---

```{r setup, include=FALSE}
library(DBI)
library(RSQLite)
con = dbConnect(SQLite(), dbname = "knock100.db")
knitr::opts_chunk$set(echo=T,warning=F,message=F,comment='',
                      connection = "con")
```

## see tables and its columns, values
```{sql}
pragma table_list;
```

```{sql}
pragma table_info(receipt);
```
```{sql}
select * from receipt limit 5;
```

```{sql}
pragma table_info(geocode);
```
```{sql}
select * from geocode limit 5;
```

```{sql}
pragma table_info(category);
```
```{sql}
select * from category limit 5;
```

```{sql}
pragma table_info(store);
```
```{sql}
select * from store limit 5;
```

```{sql}
pragma table_info(product);
```
```{sql}
select * from product limit 5;
```

```{sql}
pragma table_info(customer);
```
```{sql}
select * from customer limit 5;
```


## see duplication
```{sql}
select customer_name, postal_cd, count(1) as n
from customer
group by customer_name, postal_cd
having n!=1;
```

```{sql}
with tmp1 as(
  select customer_id, customer_name, postal_cd
    , row_number() over (partition by customer_name, postal_cd order by customer_id) as n
  from customer
)
select * from tmp1 where n!=1;
```

### reject duplication
```{sql}
select customer_name, postal_cd from customer
union
select customer_name, postal_cd from customer
```


## process to null
```{r}
dbExecute(con,"create temp table tmp01 (chr1 text, num1 int, num2 int, num3 int)")
sql=paste0("insert into tmp01 (chr1, num1, num2, num3) values "
           ,"('a',0,10,100),"
           ,"('a',1,20,200),"
           ,"('b',null,30,300),"
           ,"('b',2,null,400),"
           ,"('c',3,40,500)")
dbExecute(con,sql)
```
```{sql}
select * from tmp01;
```

### don't count null
```{sql}
select count(1), count(chr1), count(num1), count(num2), count(num3)
  , sum(chr1), sum(num1), sum(num2), sum(num3)
  , avg(chr1), avg(num1), avg(num2), avg(num3) from tmp01;
```

### rows have null
answer of alithmetic  includes null is null,  
text type column's value is 0 when its used alithmetic
```{sql}
select * from tmp01
  where chr1+num1+num2 is null;
```

### groups have null in specified column
```{sql}
select chr1 from tmp01
group by chr1
having count(1)!=count(num1);
```


## difference of tables
```{sql}
create temp table tmp02 as
select * from tmp01;
```
```{sql}
update tmp02 set num3=0 where chr1='c';
```
```{sql}
select * from tmp02;
```

```{sql}
with tmp1 as(
  select * from tmp01
  except
  select * from tmp02
)
, tmp2 as(
  select * from tmp02
  except
  select * from tmp01
)
select 'old',* from tmp1
union all
select 'new',* from tmp2
```

```{sql}
drop table tmp01;
```
```{sql}
drop table tmp02;
```



## aggregation
```{sql}
select store_cd, count(1) as n, sum(amount) as sum
  , mode(amount) as mode
  , min(amount-0) as min
  , lower_quartile(amount) as q1
  , median(amount) as med
  , upper_quartile(amount) as q3
  , max(amount-0) as max
  
  , avg(amount) as mean
  , variance(amount) as var
  , stdev(amount) as sd
  , stdev(amount)/avg(amount) as cv
from receipt
group by store_cd
```


## median absolute deviation, MAD
```{sql}
with tmp1 as(
  select median(amount) as med from receipt 
)
select median(abs(amount-med))*1.4826 as mad
from receipt,tmp1;
```


## cross aggregation by 2 categories
```{r}
dbExecute(con,"create temp table tmp01 (chr1, chr2, num1, num2, num3, day)")

x1=sample(c('a','b','c','d'),100,replace=T,prob=c(0.6,0.3,0.15,0.05))
x2=sample(c('c1','c2'),100,replace=T,prob=c(0.7,0.3))
x3=floor(runif(100,0,10))
x4=floor(runif(100,0,100))
x5=runif(100,0,1)

date_seq=seq.Date(as.Date('2020-01-01'),as.Date('2021-12-31'),by='day')
x6=sample(date_seq,100,replace=T)

sql0="insert into tmp01 (chr1, chr2, num1, num2, num3, day) values "
sql1=paste0("('",x1,"','",x2,"',",x3,",",x4,",",x5,",'",x6,"');")

sql=paste0(sql0,sql1)

for(i in sql) dbExecute(con,i)
```
```{sql}
select * from tmp01;
```

```{sql}
select chr1
, count(case when chr2='c1' then 1 else null end) as n_c1
, sum(case when chr2='c1' then num1 else 0 end) as sum_c1
, count(case when chr2='c2' then 1 else null end) as n_c2
, sum(case when chr2='c2' then num1 else 0 end) as sum_c2
from tmp01 group by chr1;
```

```{sql}
drop table tmp01;
```


## histgram on Sturges' rule
```{sql}
with tmp1 as(
 select ceil(log(count(1)/log(2))+1) as k, min(amount-0) as min, max(amount-0) as max
 from receipt
)
select bin, min(amount-0) as left, max(amount-0) as right, count(1) from(
 select amount, round((amount-min-0.0)/(max-min)* k ) as bin from receipt, tmp1)
group by bin order by bin;
```


## correlation
```{sql}
create temp table tmp0 as
  select store_cd, floor_area as area, sum(amount) as amount
  from store left join receipt using(store_cd)
  group by store_cd, area;
```
```{sql}
select * from tmp0;
```

#### Peason's correlation
```{sql}
with tmp1 as(
  select count(1) as n from tmp0
)
select (avg(area*amount)-avg(area)*avg(amount))
  /stdev(area)/stdev(amount)*tmp1.n/(tmp1.n-1) as corr
from tmp0,tmp1;
```

#### Kendall's correlation
```{sql}
with tmp1 as(
  select area, amount from tmp0
  where area is not null and amount is not null
)
, tmp2 as(
  select sign(t1.area-t2.area) as s1, sign(t1.amount-t2.amount) as s2
  from tmp1 as t1 cross join tmp1 as t2
  where t1.area!=t2.area and t1.amount!=t2.amount
)
select sum(s1*s2)/count(1) as Kendall_corr
from tmp2;
```

#### sign correlation
```{sql}
with tmp1 as(
  select median(area) as med1, median(amount) as med2
  from tmp0
)
, tmp2 as(
  select sign(area-med1) as s1, sign(amount-med2) as s2
  from tmp0, tmp1
  where area not null and amount is not null
)
select sum(s1*s2)/count(1) as sign_corr
from tmp2;
```

```{sql}
drop table tmp0;
```


## covariance, correlation matrix
```{sql}
drop table if exists table0;
```
```{sql}
create table table0 (col1,col2,col3);
```
```{sql}
insert into table0 (col1,col2,col3) values
(2,3,5),(7,11,13),(17,19,23);
```
```{sql}
select * from table0;
```
```{sql}
with tmp1 (x1,x2,x3) as(
  select * from table0
)
select * from tmp1
;
```

```{sql}
with tmp1 (x1,x2,x3) as(
  select * from table0
)
, tmp2 as(
  select count(1) as n from tmp1
)
, tmp3 (row,col,val) as(
  select 1,1,power(stdev(x1),2) from tmp1
  union all select 1,2,(avg(x1*x2)-avg(x1)*avg(x2))*tmp2.n/(tmp2.n-1) from tmp1,tmp2
  union all select 1,3,(avg(x1*x3)-avg(x1)*avg(x3))*tmp2.n/(tmp2.n-1) from tmp1,tmp2
  union all select 2,2,power(stdev(x2),2) from tmp1
  union all select 2,3,(avg(x2*x3)-avg(x2)*avg(x3))*tmp2.n/(tmp2.n-1) from tmp1,tmp2
  union all select 3,3,power(stdev(x3),2) from tmp1
)
select * from tmp3
;
```


```{sql}
with tmp1 (x1,x2,x3) as(
  select * from table0
)
, tmp2 as(
  select count(1) as n from tmp1
)
, tmp3 (row,col,val) as(
  select 1,1,1
  union all select 1,2
  ,(avg(x1*x2)-avg(x1)*avg(x2))/stdev(x1)/stdev(x2)*tmp2.n/(tmp2.n-1) from tmp1,tmp2
  union all select 1,3
  ,(avg(x1*x3)-avg(x1)*avg(x3))/stdev(x1)/stdev(x3)*tmp2.n/(tmp2.n-1) from tmp1,tmp2
  union all select 2,2,1
  union all select 2,3
  ,(avg(x2*x3)-avg(x2)*avg(x3))/stdev(x2)/stdev(x3)*tmp2.n/(tmp2.n-1) from tmp1,tmp2
  union all select 3,3,1
)
select * from tmp3
;
```

```{sql}
drop table table0;
```



## see difference
```{sql}
create temp table tmp0 as
  with tmp1 as(
    select count(1) as n0
    , avg(age) as mean0
    , stdev(age) as sd0
    from customer
  )
  , tmp2 as(
    select count(1) as na
    , avg(age) as mean_a
    , stdev(age) as sd_a
    from customer
    where gender_cd='0'
  )
  , tmp3 as(
    select count(1) as nb
    , avg(age) as mean_b
    , stdev(age) as sd_b
    from customer
    where gender_cd='1' 
  )
  select n0, mean0, sd0, na, mean_a, sd_a, nb, mean_b, sd_b
  from tmp1, tmp2, tmp3;
```
```{sql}
select * from tmp0;
```

#### Glass's delta
```{sql}
select (mean_a-mean0)/sd0 as G_delta from tmp0;
```

#### Cohen's d
```{sql}
select (mean_a-mean_b)/sqrt(((na-1)*sd_a*sd_a+(nb-1)*sd_b*sd_b)/(na+nb-2)) as Cohen_d
from tmp0;
```

```{sql}
drop table tmp0;
```



## make dummy var. from category var.
```{sql}
select customer_id
  , max(case when gender_cd='0' then 1 else 0 end) as male
  , max(case when gender_cd='1' then 1 else 0 end) as female
  , max(case when gender_cd='9' then 1 else 0 end) as unknown
from customer
group by customer_id;
```


## invalid date value(YYYYmmdd) is made to null
```{sql}
select '20000123' as str
  , date(substr('20000123',1,4)||'-'||
    substr('20000123',5,2)||'-'||
    substr('20000123',7,2)) as date;
```

```{sql}
select '20001234' as str
  , date(substr('200001234',1,4)||'-'||
    substr('20001234',5,2)||'-'||
    substr('20001234',7,2)) as date;
```

```{sql}
select sales_ymd
  , date(substr(sales_ymd,1,4)||'-'||
    substr(sales_ymd,5,2)||'-'||
    substr(sales_ymd,7,2)) as date
from receipt;
```


## invalid time value(HHMMSS) is made to null
```{sql}
select '123456' as str
  , time(substr('123456',1,2)||':'||
    substr('123456',3,2)||':'||
    substr('123456',5,2)) as time;
```

```{sql}
select '234567' as str
  , time(substr('234567',1,2)||':'||
    substr('234567',3,2)||':'||
    substr('234567',5,2)) as time;
```


## interval of date
```{sql}
select customer_id, birth_day, apply
  , julianday(apply)-julianday(birth_day) as days
  , strftime('%Y',julianday(apply))-strftime('%Y',julianday(birth_day))-0 as years
from(
  select customer_id, birth_day
    , date(substr(application_date,1,4)||'-'||
      substr(application_date,5,2)||'-'||
      substr(application_date,7,2)) as apply
  from customer);
```


## add row has 0 instead of no row in time series
```{sql}
select customer_id
  , row_number() over(partition by customer_id order by sales_ymd) as times
from receipt
where customer_id not like 'Z%'
order by times desc;
```

```{sql}
with tmp1 as(
  select distinct strftime('%Y%m',date(sales_epoch,'unixepoch')) as ym
  from receipt
)
select ym, coalesce(sum,0) from tmp1 left join(
  select strftime('%Y%m',date(sales_epoch,'unixepoch')) as ym
    , sum(amount) as sum
  from receipt where customer_id='CS040214000008'
  group by ym
) using(ym)
order by ym;
```


## standardization, normalization
```{sql}
with tmp1 as(
  select sales_ymd, sum(quantity) as qty, sum(amount) as amt
  from receipt group by sales_ymd
)
, tmp2 as(
  select avg(qty) as q_mean, avg(qty*qty) q_sq
    , min(qty) as q_min, max(qty) as q_max
    , avg(amt) as a_mean, avg(amt*amt) a_sq
    , min(amt) as a_min, max(amt) as a_max
  from tmp1
)
select sales_ymd
  , (qty-q_mean)/sqrt(q_sq-q_mean*q_mean) as q_z
  , (qty-q_min-0.0)/(q_max-q_min) as q_0to1
  , (amt-a_mean)/sqrt(a_sq-a_mean*a_mean) as a_z
  , (amt-a_min-0.0)/(a_max-a_min) as a_0to1
from tmp1,tmp2;

```


## difference to previous period, rate to previous period
```{sql}
with tmp1 as(
  select sales_ymd, sum(quantity) as qty, sum(amount) as amt
  from receipt group by sales_ymd
)
, tmp2 as(
  select sales_ymd
    , qty, lag(qty,1) over(order by sales_ymd) as qty_lag1
    , lag(qty,2) over(order by sales_ymd) as qty_lag2
    , amt, lag(amt,1) over(order by sales_ymd) as amt_lag1
    , lag(amt,2) over(order by sales_ymd) as amt_lag2
  from tmp1
)
select sales_ymd, qty, qty_lag1
  , qty-qty_lag1 as q_diff1, qty*1.0/qty_lag1 as q_rate1, qty_lag2
  , amt, amt_lag1
  , amt-amt_lag1 as a_diff1, amt*1.0/amt_lag1 as a_rate1, amt_lag2
from tmp2;
```


## cumulative sum, moving average
```{sql}
with tmp1 as(
  select sales_ymd, sum(quantity) as qty, sum(amount) as amt
  from receipt group by sales_ymd
)
select sales_ymd
  , qty, sum(qty) over(order by sales_ymd) as cum_qty
  , avg(qty) over(order by sales_ymd) as cum_avg_qty
  , avg(qty) over(order by sales_ymd rows between 2 preceding and current row) as mov_avg_qty
  , amt, sum(amt) over(order by sales_ymd) as cum_amt
  , avg(amt) over(order by sales_ymd) as cum_avg_amt
  , avg(amt) over(order by sales_ymd rows between 2 preceding and current row) as mov_avg_amt
from tmp1;
```



## recursive CTE

### make number sequence
```{sql}
with tmp1 as(
  select 1 as num
  union all
  select num+1 from tmp1
  where num<10
  --limit 10
)
select * from tmp1;
```

### make date sequence
```{sql}
with tmp1 as(
  select '2024-01-01' as date
  union all
  select date(julianday(date)+1) from tmp1
  where date<'2024-01-10'
)
select * from tmp1;
```

### Fibonacci sequence
```{sql}
with tmp1 as(
  select 1 as n, 0 as val, 1 as pre
  union all
  select n+1, val+pre, val
  from tmp1
  where n<10
)
select n, val from tmp1
;
```


### prime number
```{sql}
drop table if exists prime;
```
```{sql}
create table prime (p int);
```
```{sql}
with tmp1 as (
  select 2 as n
  union all
  select n+1 from tmp1 where n<100000
),
tmp2 as (
  select t1.n from tmp1 as t1
  where not exists (
    select 1 from tmp1 as t2
    where t2.n<t1.n and t1.n%t2.n=0
    --where t2.n*t2.n<=t1.n and t1.n%t2.n=0
  )
)
insert into prime (p)
  select n from tmp2;
```

```{sql}
select * from prime limit 10;
```
```{sql}
select count(1) from prime;
```

```{sql}
drop table prime;
```

### prime number using rowid
```{sql}
drop table if exists prime;
```
```{sql}
create table prime (p int);
```
```{sql}
drop table if exists dum;
```
```{sql}
create table dum (dum int);
```
```{sql}
with tmp1 as (
  select 1 as n
  union all
  select n+1 from tmp1 where n<100000
)
insert into dum (dum)
  select null from tmp1;
```
```{sql}
with tmp1 as (
  select t1.rowid as n from dum as t1
  where t1.rowid>1
  and not exists (
    select 1 from dum as t2
    where t2.rowid>1 and t2.rowid<t1.rowid and t1.rowid%t2.rowid=0
  )
)
insert into prime (p)
  select n from tmp1;
```

```{sql}
select * from prime limit 10;
```
```{sql}
select count(1) from prime;
```

```{sql}
drop table dum;
```

### prime factorization
```{sql}
with tmp1(n,p) as(
  select 1111111111, 1 -- n, 1
  union all
  select n/(select min(p) from prime where n%p=0)
  , (select min(p) from prime where n%p=0)
  from tmp1 where n>1
)
select * from tmp1;
```

#### prime factorization using conditional union
```{sql}
with tmp1(n, p) as(
  select 1111111111, 2 -- n, 2 minimum prime

  union all
  select n/p, p
  from tmp1 where n%p=0 --case1 keep dividing n by p 

  union all
  select n, (select p from prime where p>tmp1.p limit 1)  
  from tmp1 where n%p<>0 --case2 get next minimum prime
)
select * from tmp1 where n%p=0;
```

```{sql}
drop table prime;
```


### hierarchy level
```{sql}
create table table0 (id,child,parent);
```
```{sql}
insert into table0 (id,child,parent) values
(1,'a',null)
,(2,'aa','a')
,(3,'ab','a')
,(4,'ac','a')
,(5,'aaa','aa')
,(6,'aab','aa')
,(7,'aba','ab')
,(8,'abb','ab')
,(9,'aca','ac')
,(10,'aaaa','aaa')
,(11,'aaab','aaa')
,(12,'acac','aca')
;
```
```{sql}
select * from table0;
```
 
```{sql}
with tmp1 as(
  select id,child as name,parent,1 as lv
  from table0
  where parent is null --parent of root is null
  union all
  select t0.id,t0.child,t0.parent,tmp1.lv+1
  from table0 as t0
  inner join tmp1 on t0.parent=tmp1.name
)
select * from tmp1;
```
 
```{sql}
drop table table0;
```

### back to root
```{sql}
create table table0 (id,name,parent);
```
```{sql}
insert into table0 (id,name,parent) values
  (1, 'Alice', NULL)
, (2, 'Bob', 1)
, (3, 'Charlie', 1)
, (4, 'David', 2)
, (5, 'Eve', 2)
, (6, 'Frank', 3)
, (7, 'Grace', 5)
, (8, 'Hannah', 4)
, (9, 'Ivy', 4)
, (10, 'Jack', 5)
;  
```
```{sql}
select * from table0;
```

```{sql}
with tmp1 as(
  select id, parent, name, 0 as era
  from table0
  where id=7 -- specify child
  union all
  select table0.id, table0.parent, table0.name, tmp1.era+1
  from table0
  inner join tmp1 on table0.id=tmp1.parent
)
select * from tmp1
order by era
;
```

```{sql}
drop table table0;
```

### count components
```{sql}
create table table0 (id,child,qty);
```
```{sql}
insert into table0 (id,child,qty) values
  ('a', 'b', 2)
, ('a', 'c', 1)
, ('b', 'd', 4)
, ('b', 'e', 2)
, ('c', 'f', 3)
, ('c', 'g', 2)
;
```
```{sql}
select * from table0;
```

```{sql}
with tmp1 as (
  select id, child, qty, qty as acm
  from table0
  where id='a' -- specify parent
  union all
  select table0.id, table0.child, table0.qty
  , tmp1.acm*table0.qty as acm
  from table0
  inner join tmp1 on table0.id=tmp1.child
)
select child
, sum(acm) as total
from tmp1
group by child
;
```

```{sql}
drop table table0;
```



### fill null by previous row value 
```{sql}
create table table0 (id,val1,val2);
```
```{sql}
insert into table0 (id,val1,val2) values
 (2, 20, 200)
,(1, 10, 100)
,(NULL, 200, NULL)
,(4, 40, 400)
,(NULL, NULL, 4000)
,(NULL, NULL, 40000)
,(3, 30, 300)
;
```
```{sql}
select * from table0;
```

```{sql}
with recursive tmp1 as(
 select rowid, id, val1, val2 from table0
 where id is not null
 union all
 select table0.rowid
 ,coalesce(table0.id, tmp1.id)
 ,table0.val1, table0.val2
 from table0
 inner join tmp1 on table0.rowid=tmp1.rowid+1
 where table0.id is null
)
select id, val1, val2 from tmp1
order by rowid
;
```

```{sql}
drop table table0;
```



### minimum depth on graph
```{sql}
create table table0 (node1,node2);
```
```{sql}
insert into table0 (node1,node2) values
('a','b')
,('a','c')
,('b','d')
,('c','e')
,('c','f')
,('e','g')
,('f','h')
,('g','i')
,('i','j')
,('c','j')
;
```
```{sql}
select * from table0;
```

```{sql}
with tmp1 AS (
  select node1, node2, 1 as depth, node1||','||node2 as visited
  from table0
  where node1='a' --start node
  union all
  select t0.node1, t0.node2, tmp1.depth+1
  , tmp1.visited||','||t0.node2 as visited
  from table0 as t0
  inner join tmp1 on t0.node1=tmp1.node2
  where tmp1.depth<10 -- Limit search depth
    and ','||tmp1.visited||',' not like '%,'||t0.node2||',%'
)
select node2 AS goal, min(depth) as depth
, visited from tmp1
group by goal order by depth;
```

```{sql}
drop table table0;
```

### minimum distance on graph
```{sql}
create table table0 (node1,node2,weight);
```
```{sql}
insert into table0 (node1,node2,weight) values
('a','b',1)
,('a','c',2)
,('b','d',3)
,('c','e',2)
,('c','f',1)
,('e','g',2)
,('f','h',3)
,('g','i',2)
,('i','j',1)
,('c','j',2)
;
```
```{sql}
select * from table0;
```

```{sql}
with tmp1 as(
  select node1, node2, 1 as depth, weight
  , '>'||node2 as path
  , ','||node1||','||node2 as visited
  from table0
  where node1='a' --start node
  union all
  select t0.node1, t0.node2
  , tmp1.depth+1
  , tmp1.weight+t0.weight
  , tmp1.path||' >'||t0.node2 as path
  , tmp1.visited||','||t0.node2 as visited
  from table0 as t0
  inner join tmp1 on t0.node1=tmp1.node2
  where tmp1.weight<100 --max search depth
    and ','||tmp1.visited||','  not like '%,'||t0.node2||',%'
)
select node2 as goal
, depth
, min(weight) as distance
, path
from tmp1
group by goal order by distance
;
```

```{sql}
drop table table0;
```


### see cycle on graph
```{sql}
create table table0 (node1, node2);
```
```{sql}
insert into table0 (node1, node2) values
  ('a', 'b')
, ('b', 'c')
, ('c', 'd')
, ('d', 'b') -- cycle b >c >d >b
, ('e', 'f')
, ('f', 'g')
, ('g', 'e') -- cycle e >f >g >e
;
```
```{sql}
select * from table0;
```

```{sql}
with tmp1 as(
  select node1 ||' >'|| node2 as path
  , node2 as current
  , 1 as depth
  from table0
  union all
  select path ||' >'|| node2, node2, depth+1
  from tmp1
  inner join table0 on tmp1.current=table0.node1
  where depth<30 --less than #node
)
select path from tmp1
where path like current||' >% >'||current
  and path not like current||' >% >'||current||' >%'
;
```

```{sql}
drop table table0;
```


### split text by some characters
```{sql}
create table table0 (id,txt);
```
```{sql}
insert into table0 (id,txt) values
  (1,'abcd efgh')
, (2,'ijk lmn opq')
, (3,'rstu vwxyz')
;	
```
```{sql}
select * from table0;
```

```{sql}
with tmp1 as(
  select id, 1 as pos, substr(txt,1,2) as chrs --by 2 characters
  from table0
  union all
  select tmp1.id, pos+1, substr(table0.txt,pos+1,2) --by 2 characters
  from tmp1 join table0 on tmp1.id=table0.id
  where pos<length(table0.txt)
)
select * from tmp1
where not length(replace(chrs,' ',''))<2
order by id, pos
;
```

```{sql}
with tmp1 as(
  select id, 1 as pos, substr(txt,1,3) as chrs --by 3 characters
  from table0
  union all
  select tmp1.id, pos+1, substr(table0.txt,pos+1,3) --by 3 characters
  from tmp1 join table0 on tmp1.id=table0.id
  where pos<length(table0.txt)
)
select * from tmp1
where not length(replace(chrs,' ',''))<3
order by id, pos
;
```

```{sql}
drop table table0;
```


## node and edge graph
```{sql}
create temp table edges0 as
  with tmp1 as (
      select 1 as num
      union all
      select num+1 from tmp1 where num<300
  )
  select abs(random())%10+1 as org, abs(random())%10+1 as dst from tmp1;
```
```{sql}
create temp table edges as
  select * from edges0 where org!=dst;
```
```{sql}
select * from edges;
```

```{sql}
with tmp1 as(
  select org, count(1) as out_deg from edges group by org
)
, tmp2 as(
  select dst, count(1) as in_deg from edges group by dst
)
, tmp3 as(
  select org, dst, count(1) as weight from edges group by org,dst
)
select org, dst, weight, out_deg, weight*1.0/out_deg as out_prop
  , in_deg, weight*1.0/in_deg as in_prop
from tmp3
inner join tmp1 using(org)
inner join tmp2 using(dst);
```

```{sql}
drop table edges;
```

```{sql}
create temp table edges as
  select case when org<dst then org else dst end as node1
    , case when org>dst then org else dst end as node2
  from edges0
```
```{sql}
select * from edges;
```

```{sql}
with tmp1 as(
  select node1, count(1) as deg1 from edges group by node1
)
, tmp2 as(
  select node2, count(1) as deg2 from edges group by node2
)
, tmp3 as(
  select node1, node2, count(1) as weight from edges group by node1, node2
)
select node1, node2, weight, deg1+deg2, weight*1.0/(deg1+deg2) as prop
from tmp3
inner join tmp1 using(node1)
inner join tmp2 using(node2);
```

```{sql}
drop table edges;
```
```{sql}
drop table edges0;
```


## basket analysis, collaborative filtering
```{sql}
create temp table basket0 as
  with tmp1 as (
      select 1 as num
      union all
      select num+1 from tmp1 where num<100
  )
  select abs(random())%10+1 as grp, abs(random())%20+1 as item from tmp1;
```

#### It does not has order
```{sql}
create temp table basket as
  with tmp1 as(
    select grp, item
    from basket0
  )
  , tmp2 as(
    select t1.grp as grp, t1.item as item1, t2.item as item2
    from tmp1 as t1
    inner join tmp1 as t2 on t1.grp=t2.grp and t1.item<t2.item
    group by t1.grp, item1, item2
  )
  select grp, item1, item2 from tmp2;
```
```{sql}
select * from basket order by grp;
```

```{sql}
select item1, item2, count(1) as frq from basket group by item1, item2
order by item1, item2; 
```

```{sql}
drop table basket;
```


#### It has order, make trigram
```{sql}
create temp table basket as
  with tmp1 as(
    select distinct grp, item from basket0
  )
  select grp, item
    , row_number() over(partition by grp order by random()) as ord
  from tmp1
```

```{sql}
select grp, ord
  , lag(item,1) over(partition by grp) as before
  , item
  , lead(item,1) over(partition by grp) as after
from basket;
```

```{sql}
drop table basket;
```
```{sql}
drop table basket0;
```


## process of categorical variable
```{sql}
select gender_cd, count(1), count(1)*1.0/max(rowid) as proportion
from customer group by gender_cd;
```

```{sql}
select gender_cd, age/10*10 as ages, count(1)
from customer group by gender_cd, ages;
```

```{sql}
select customer_id, gender_cd
  , case when age-0<20 then 1 else 0 end young
  , case when age-0 between 20 and 64 then 1 else 0 end adult
  , case when age-0>64 then 1 else 0 end senior
from customer limit 10;
```


```{sql}
create temp table cross as
  with tmp1 as(
    select gender_cd
    , sum(case when age-0<20 then 1 else 0 end) young
    , sum(case when age-0 between 20 and 64 then 1 else 0 end) adult
    , sum(case when age-0>64 then 1 else 0 end) senior
    , count(1) subtotal
    from customer group by gender_cd
  )
  select * from tmp1
```

```{sql}
select * from cross
union all
select 'subtotal', sum(young), sum(adult), sum(senior), sum(subtotal)
from cross;
```

```{sql}
select gender_cd
  , young*1.0/subtotal as prop_young
  , adult*1.0/subtotal as prop_adult
  , senior*1.0/subtotal as prop_senior
from cross
union all
select 'total'
  , sum(young)*1.0/sum(subtotal)
  , sum(adult)*1.0/sum(subtotal)
  , sum(senior)*1.0/sum(subtotal)
from cross;
```
```{sql}
with tmp1 as(
  select 'proportion'
  , sum(young)*1.0/sum(subtotal) as prop_young
  , sum(adult)*1.0/sum(subtotal) as prop_adult
  , sum(senior)*1.0/sum(subtotal) as prop_senior
  from cross
)
select gender_cd
  , young*1.0/subtotal/tmp1.prop_young as lift_young
  , adult*1.0/subtotal/tmp1.prop_adult as lift_adult
  , senior*1.0/subtotal/tmp1.prop_senior as lift_senior
from cross,tmp1;
```

```{sql}
drop table cross;
```



## sampling
#### extract a certain number
```{sql}
select row_number() over() as id,* from(
  select customer_id, gender_cd, age from customer
  where customer_id not like 'Z%'
  order by random() limit 10
)
```

#### stratify and extract a certain number
```{sql}
select row_number() over() as id,* from(
  select customer_id, gender_cd, age from customer
  where customer_id not like 'Z%' and gender_cd='0'
  order by random() limit 10
)
union all
select row_number() over() as id,* from(
  select customer_id, gender_cd, age from customer
  where customer_id not like 'Z%' and gender_cd='1'
  order by random() limit 10
)
```

#### divide data to learn or test with 8:2 for ML
```{sql}
select row_number() over() as id, customer_id
  , case when abs(random())%10<8 then 'learn' else 'test' end as grp
from customer
where customer_id not like 'Z%';
```

#### divide to 2 groups 
```{sql}
select row_number() over() as id, customer_id, gender_cd, age
  , case when abs(random())%2<1 then 'A' else 'B' end as grp
from customer
where customer_id not like 'Z%';
```

#### divide to 4 groups 
```{sql}
create temp table tmp1 as
  select row_number() over() as id, customer_id, gender_cd, age
    , case when abs(random())%4<1 then 'A'
      when abs(random())%3<1 then 'B'
      when abs(random())%2<1 then 'C' else 'D' end as grp
  from customer
  where customer_id not like 'Z%';
```

```{sql}
select grp, gender_cd, count(1), avg(age) from tmp1 group by grp, gender_cd;
```

```{sql}
drop table tmp1
```


